---
title: "EDA-Project-R"
output: html_document
---

# NTSB Aviation Accident Database Exploration by Hamza Bendemra

## Introduction
This dataset was acquired from [Kaggle](https://www.kaggle.com/) and provides data from US [NTSB Aviation Accident Database](https://www.ntsb.gov/_layouts/ntsb.aviation/index.aspx). The clean dataset provided by Kaggle can be [found here](https://www.kaggle.com/khsamaha/aviation-accident-database-synopses). 

### Libraries
First, let's import the libraries of interested for this EDA.

```{r}
library(gridExtra)
library(ggplot2)
library(ggthemes)
library(tidyr)
library(dplyr)
library(lubridate)
library(reshape2)
library(GGally)
library(scales)
library(memisc)
library(RColorBrewer)
#theme_set(theme_minimal(10)) 
```

## Importing the native CSV file into a daframe
Then, lets' import the dataset in an R dataframe 'pf', and have an initial look at it.

```{r}
pf <- read.csv('AviationData.csv')
```



## Plots and Analysis 
In this section, I'll be generate plots and discussion in the results as part of an analysis. Firstly, we'll perform a univaritate exploration, then move to bivariate analysis, and finally perform multivariate analysis on the dataset.

### Univariate Plots and Analysis

### Bivariate Plots and Analysis

### Multivariate Plots and Analysis

## Final Plots and Summary
In this section, I selected three plots from my analysis that show different trends worth noting.

## Reflection
In this section, I will reflect on the struggles, successes, and ideas for future exploration on the data set. 

## Document your Data


