---
title: "NTSB Aviation Accident Database Exploration"
author: "Hamza Bendemra"
output: github_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(gridExtra)
library(datasets)
library(ggplot2)
library(ggthemes)
library(tidyr)
library(dplyr)
library(lubridate)
library(reshape2)
library(GGally)
library(scales)
library(memisc)
library(RColorBrewer)
library(scales)
library(forcats)
library(Hmisc)
library(maps)
library(mapproj)
rm(list = ls()) # clear workspace
# colours for the histograms
barfill <- "#4271AE"
barlines <- "#1F3552"
#colour <- "black"
fill <- "#56B4E9"
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

## Introduction
In this repo, I perform an Exploratory Data Analysis (EDA) on a Kaggle dataset.
The dataset was sourced from The NTSB Aviation Accident Database which 
contains information from 1962 and later about civil aviation accidents and 
selected incidents within the United States and in international waters.

* Method: Exploratory Data Analysis (EDA)
* Program: R programming language, RStudio
* Topic: Commercial Aviation
* Sources: NTSB Aviation Accident Database

This EDA does not aim to answer a particular question, but rather explore the
data and get an initial look at relationships between the various variables
present in our dataset.

# Introduction
Our dataset features 81,013 observations of 31 variables (shown below) which 
are related to aviation accidents recorded. Variables provide information on a 
variety of topics including date and location of observations, model and type of 
aircraft, information on the sustained injuries to passengers and to the 
aircraft, and the reported weather conditions at the time.

```{r message=FALSE, warning=FALSE, Load_the_Data}
pf <- read.csv('AviationData.csv') # Load the Data
```

```{r message=FALSE, warning=FALSE}
dim(pf)
```

```{r message=FALSE, warning=FALSE}
names(pf)
```

The variable types include factorial, numeric, and integer variables.

```{r message=FALSE, warning=FALSE}
str(pf)
```

# Data Wrangling
Since this an NTSB database from the United States, it is highly likely 
that the majority of the reported observations are from the US. Let's check 
that by calculating the percentage of reported accidents in this database 
from the United States.

```{r message=FALSE, warning=FALSE}
dim(subset(pf, Country == "United States")) / dim(pf)
```

As expected, the majority of accidents (over 94%) in this databased are from 
the US. Hence, we will be focusing on the accidents that took place in the US 
in this analysis.

```{r message=FALSE, warning=FALSE}
pf_us = subset(pf, Country == "United States")
```

As shown below, the new dataframe now features 76,188 observations.

```{r message=FALSE, warning=FALSE}
dim(pf_us)
```

Let's separate the listed location names (city, state) into two separate
variables: one for the city and one for the state, for each observation.

```{r}
pf_us <- separate(pf_us, Location, c("City","State"), sep=",")
```

We will further tailor our dataframe for this analysis by focusing on a 
specific number of variables of interest, out of the 31 available variables.

Let's focus on: 
(1) the date, 
(2) the total fatal injuries, 
(3) the engine type,
(4) the weather condition,
(5) the broad phase of flight,
(6) Latitude, 
(7) Longitude, 
(8) City,
(9) State.


```{r message=FALSE, warning=FALSE}
vars_wanted <- c("Event.Date", #1
                 "Total.Fatal.Injuries", #2
                 "Engine.Type", #3
                 "Weather.Condition", #4
                 "Broad.Phase.of.Flight", #5
                 "Latitude", #6
                 "Longitude", #7
                 "City", #8
                 "State",
                 "Aircraft.Category"
                 )
pf_study <- pf_us[vars_wanted]
```

Let's check that our new dataframe 'pf_study' features all 5 variables.

```{r message=FALSE, warning=FALSE}
names(pf_study)
```

```{r message=FALSE, warning=FALSE}
dim(pf_study)
```

As shown above, the dataframe we will be using for this study features 
76,188 observations of 5 variables. Below is a summary of the dataframe.

Looking at the summary output we can get a quick overview of the distribution 
for each of our variables.

```{r message=FALSE, warning=FALSE}
summary(pf_study)
```

The variable related to the observation's event date will need some further 
data wrangling to be able to extract meanigful information to draw conclusions 
from. For instance, it would be useful to break the observation's event date 
by day, month, and year to investigate if there are any correlations between 
number of accidents/incidents and particular periods within a year.

```{r message=FALSE, warning=FALSE}
pf_study$Event.Date <- as.Date(pf_study$Event.Date)
pf_study$Event.Year <- format(pf_study$Event.Date,"%Y")
pf_study$Event.Year <- as.integer(pf_study$Event.Year)
my_breaks = c(1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010)
pf_study$Event.Month <- format(pf_study$Event.Date,"%b")
pf_study$Event.Month <- factor(pf_study$Event.Month, 
                               levels = c("Jan", "Feb", "Mar", 
                                          "Apr", "May", "Jun",
                                          "Jul", "Aug", "Sep",
                                          "Oct", "Nov", "Dec"))
pf_study$Event.Weekday <- format(pf_study$Event.Date,"%a")
pf_study$Event.Weekday <- factor(pf_study$Event.Weekday, 
                                 levels = c("Mon", "Tue", "Wed", 
                                            "Thu", "Fri", "Sat", "Sun"))
```

Also, looking at the Total Fatalities, we can see that the 3rd Qu. is 1.000,
the Max is 265. Clearly, this distribution will be heavily skewed. There is 
also some entries with missing data values for the Fatalities count.
We will assume that those entries corresponds to no fatalities (i.e. 0).

A better way of displaying this data is by grouping number of fatalities in 
buckets, as shown below. This will give us a better reprsentation of the 
distribution of fatalities across the observations in the dataset.

```{r message=FALSE, warning=FALSE}
pf_study$Total.Fatal.Injuries[which(is.na(pf_study$Total.Fatal.Injuries))] <- 0
pf_study$Total.Fatal.Injuries[pf_study$Total.Fatal.Injuries==""] <- 0
```

```{r message=FALSE, warning=FALSE}
pf_study$Total.Fatal.Injuries.Brackets <- pf_study$Total.Fatal.Injuries
pf_study$Total.Fatal.Injuries.Brackets = cut(pf_study$Total.Fatal.Injuries.Brackets, 
                                   breaks = c(0,1, 2, 4, 3, 5, 10, 20, 50, 
                                              100, 150, 200, 250, 300),
                                   right = FALSE,
                                   include.lowest = TRUE)
table(pf_study$Total.Fatal.Injuries.Brackets, useNA='ifany')
```

In terms of the weather conditions, let's replace blank entries in the weather 
conditions with "unknown". Similarly, let's replace blank entries in the Phases of Flight with "unknown".

```{r message=FALSE, warning=FALSE}
pf_study$Weather.Condition <- as.character(pf_study$Weather.Condition)
pf_study$Weather.Condition[is.na(pf_study$Weather.Condition)] <- "UNK"
pf_study$Weather.Condition <- factor(pf_study$Weather.Condition, 
                                     levels = c("VMC", "IMC", "UNK"))
pf_study$Broad.Phase.of.Flight <- as.character(pf_study$Broad.Phase.of.Flight)
pf_study$Broad.Phase.of.Flight[pf_study$Broad.Phase.of.Flight == ""] <- NA
```

Let's create dataframe for the maps we'll be creating later on.

```{r}
pf_study_map <- subset(pf_study, !is.na(Longitude))
pf_study_map_Loc <- pf_study_map[
  (duplicated(pf_study_map[c("Latitude","Longitude")]) |
    duplicated(pf_study_map[c("Latitude","Longitude")], fromLast = TRUE)), ]

pf_study_map_Loc <- subset(pf_study_map_Loc, !is.na(Longitude))
pf_study_map_Loc$Coordinates<- paste(pf_study_map_Loc$Latitude, 
                                     pf_study_map_Loc$Longitude,sep = "/")
```

# Univariate Plots Section
In this section, we will create univariate plots for variables of interest. 

## Event Date
### By Year
Now, let's plots frequency histograms for the year, month, and weekday of 
accidents/incidents in the dataset.

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, aes(x = Event.Year)) +
  geom_histogram(stat="count", alpha = 0.5,
                 colour = barlines, fill = barfill, binwidth = 1) +
  scale_x_continuous(name = "Year",
                     breaks = seq(1900, 2100, 5)) +
  scale_y_continuous(name = "Count") +
  ggtitle("Year of Recorded Accident/Incident (1948-2017)")
```

It seems the majority of the observations in the dataframe are from after the 
early 1980s onwards. Let's perform a quick check.

```{r message=FALSE, warning=FALSE}
summary(pf_study$Event.Year)
```

Let's generate a plot from 1980 to 2017.

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, aes(x = Event.Year)) +
  geom_histogram(stat="count", alpha = 0.5,
                 colour = barlines, fill = barfill, binwidth = 1) +
  scale_x_continuous(name = "Year",
                     limits = c(1982, 2018),
                     breaks = seq(1982, 2018, 2)) +
  scale_y_continuous(name = "Count", 
                     limits = c(0, 3500),
                     breaks = seq(0, 3500, 500)) +
  ggtitle("Year of Recorded Accident/Incident (1982-2017)")
```

According to our plot, the number of accidents/incidents has substantially 
decreseased between 1982 and 2017 from approx. 3400 observations to 
approx. 1600 observations.

### By Month of the Year
Next, let's look at observations distribution by months of the year.

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, aes(x = Event.Month)) +
  geom_histogram(stat="count", alpha = 0.5,
                 colour = barlines, fill = barfill) +
  scale_y_continuous(name = "Count") +
  ggtitle("Month of Recorded Accident/Incident (1948-2017)")
```

According to our plot, the highest frequency of accidents/incidents in a given 
year take place during northern hemisphere summer time (Jun-Jul-Aug).

### By Day of the Week
And finally, let's look at the observations distrubution by day of the week. 

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, aes(x = Event.Weekday)) +
  geom_histogram(stat="count", alpha = 0.5,
                 colour = barlines, fill = barfill) +
  scale_y_continuous(name = "Count") +
  ggtitle("Weekday of Recorded Accident/Incident (1948-2017)")
```

According to our plot, the highest frequency of accidents/incidents in a given 
week take place during the weekend (Sat-Sun).

## Total Fatal Injuries
Our next variable of interest relates to the Total Fatal Injuries for each 
observation in the dataset. This is quantified by the number of people injured 
for each recorded observation.

```{r message=FALSE, warning=FALSE}
grid.arrange(
ggplot(pf_study, aes(x = Total.Fatal.Injuries)) +
  geom_histogram(stat="count", alpha = 1,
                 colour = barlines, fill = barfill) +
  scale_y_log10(name = "Count (Log10)",labels = comma),

ggplot(pf_study, aes(x = Total.Fatal.Injuries)) +
  geom_line(stat="count", alpha = 1,
                 colour = barlines) +
  scale_y_log10(name = "Count (Log10)",labels = comma)  , nrow =1)
```

We can see that the majority of recorded accidents/incidents have a low number 
of fatalities while some outliers are displaying large numbers of fatalities.

Let's group the number of fatalities in buckets as shown in the plot below.

```{r message=FALSE, warning=FALSE}
ggplot(data = subset(pf_study, !is.na(Total.Fatal.Injuries.Brackets)), 
       aes(x = Total.Fatal.Injuries.Brackets)) +
  geom_histogram(stat="count", alpha = 0.5,
                 colour = barlines, fill = barfill, binwidth = 1) +
  scale_y_log10(name = "Count (Log10)",labels = comma) +
  scale_x_discrete(labels = c('1','2','3', '4', '5', 
                              '6-10', '11-20', '21-50', '51-100',
                              '101-150', '151-200', '201-250', '251-300')) +
  ggtitle("Number of Fatalities (1948-2017)")
```

As previously stated, the charts above show that the bulk of accidents 
resulting in fatalities of less than 20.

## Engine Types
Next, we look at the engine types recorded in our observations. I've abbreviate
engine type names to improve labelling of the x-axis. Note the use of the Log10
scale for the y-axis in the plot below.

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, aes(x = Engine.Type)) +
  geom_histogram(stat="count", alpha = 0.5,
                 colour = barlines, fill = barfill) +
  scale_y_log10(name = "Count (Log10)", labels = comma) +
  scale_x_discrete(labels = abbreviate) +
  ggtitle("Engine Type in Observations (1948-2017)")
```

According to our charts, the bulk of engine types in the reported
accidents/incidents are Reciprocal engine types.

## Weather Conditions
Next, we look at the weather conditions recorded in our observations. 
FYI, VMC stands for visual meteorological conditions and IMC stands for 
instrument meteorological conditions. VMC means that conditions are such that 
pilots have sufficient visibility to fly the aircraft maintaining visual 
separation from terrain and other aircraft. 
IMC means weather conditions require pilots to fly primarily by reference to 
instruments.

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, aes(x = Weather.Condition)) +
  geom_histogram(stat="count", alpha = 0.5,
                 colour = barlines, fill = barfill) +
  scale_y_continuous(name = "Count", labels = comma) +
  ggtitle("Weather Conditions of Accidents/Incidents (1948-2017)")
```

According to our plot, the bulk of accidents/incidents take place during 
VMC weather conditions.

## Broad Phases of Flight
Next, we look at the weather conditions recorded in our observations.

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, aes(x = Broad.Phase.of.Flight)) +
  geom_histogram(stat="count", alpha = 0.5,
                 colour = barlines, fill = barfill) +
  scale_y_continuous(name = "Count", labels = comma) +
  scale_x_discrete(labels = abbreviate) +
  ggtitle("Broad Phase of Flight of Observations (1982-2017)")
```

According to our plot, the bulk of accidents/incidents take place during 
landing or take-off.

# Bivariate Plots Section
Let's look at the relationship between pairs of variables that could show
interesting relationship. First a recap of available variables in our dataframe.

```{r message=FALSE, warning=FALSE}
names(pf_study)
```

## Engine Types and Total Fatal Injuries
Let's look at the relationship of Engine Types and Total Fatal Injuries.

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, aes(x = Engine.Type, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.3) +
  scale_x_discrete(labels = abbreviate) +
  ggtitle("Total Fatal Injuries vs Engine Type (1948-2017)")
```

We can see that given that the bulk of the distribution has a total fatal 
injuries under 10, let's zoom in on that portion of the data.

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, Total.Fatal.Injuries < 10), 
       aes(x = Engine.Type, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.2) +
  scale_x_discrete(labels = abbreviate) +
  #scale_y_log10(labels = comma) +
  ggtitle("Total Fatal Injuries (<10) vs Engine Type (1948-2017)")
```

According to the plot, the bulk of the data for fatalities under 20 is with the
engine type Reciprocated. The first plot shows that the Tubo-Fan engine has
more outliers with higher number of fatalities than other engins. This is likely
due to the use of Turbo-Fan engines use on some large commercial aircraft.

## Weather Conditions and Total Fatal Injuries
Let's look at the relationship of Weather Conditions and Total Fatal Injuries.

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, !is.na(Weather.Condition) & Weather.Condition != "UNK"), 
       aes(x = Weather.Condition, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.3) +
  ggtitle("Weather Conditions vs Weather Condition (1948-2017)")
```

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, Total.Fatal.Injuries < 10 &
                !is.na(Weather.Condition) & Weather.Condition != "UNK"), 
       aes(x = Weather.Condition, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.2) +
  scale_x_discrete(labels = abbreviate) +
  ggtitle("Total Fatal Injuries (<10) vs Weather Condition (1948-2017)")
```

Weather conditions do not show a particularly strong relationship with total
fatal injuries. The bulk of the distribution is associate with VMC weather
conditions. However, that is likely to the fact that the vast majority of 
flights are flown in VMC conditions.

## Phase of Flight and Total Fatal Injuries
Let's look at the relationship of Phase of Flight and Total Fatal Injuries

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, 
       aes(x = Broad.Phase.of.Flight, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.2) +
  scale_x_discrete(labels = abbreviate) +
  ggtitle("Total Fatal Injuries vs Broad Phase of Flight (1948-2017)")
```

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, Total.Fatal.Injuries < 10 & Broad.Phase.of.Flight!= "UNKNOWN"),  
       aes(x = Broad.Phase.of.Flight, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.2) +
  scale_x_discrete(labels = abbreviate) +
  ggtitle("Total Fatal Injuries (<10) vs Broad Phase of Flight (1948-2017)")
```

The plots show that Take-Off and Apprach are associated with outliers with 
high number of fatalities. These two phases of flight are often referred to as
"critical phases of flight" for that particular reason. They're high risk.

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, Total.Fatal.Injuries < 10 & Broad.Phase.of.Flight!= "UNKNOWN"),   
       aes(x = Broad.Phase.of.Flight, y = Total.Fatal.Injuries)) +
  geom_boxplot(alpha = 0.2) +
  scale_x_discrete(labels = abbreviate) +
  ggtitle("Total Fatal Injuries (<10) vs Broad Phase of Flight (1948-2017)")
```

This boxplot further emphasises the presence of high outliers for take-off
and approaches phases of flight. It also indicates that outliers are present
for all broad phases of flight.

## Event Date and Total Fatal Injuries
Let's look at the relationship of Event Date and Total Fatal Injuries

### By Month of the Year

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, 
       aes(x = Event.Month, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.2) +
  ggtitle("Total Fatal Injuries vs Event Month (1948-2017)")
```

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, Total.Fatal.Injuries < 10),  
       aes(x = Event.Month, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.2) +
  ggtitle("Total Fatal Injuries (<10) vs Event Month (1948-2017)")
```

The distribution is fairly distributed across all months of the year.

### By Day of the Week

```{r message=FALSE, warning=FALSE}
ggplot(pf_study, 
       aes(x = Event.Weekday, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.3) +
  ggtitle("Total Fatal Injuries vs Event Weekday (1948-2017)")
```

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, Total.Fatal.Injuries < 10),  
       aes(x = Event.Weekday, y = Total.Fatal.Injuries)) +
  geom_jitter(alpha = 0.3) +
  ggtitle("Total Fatal Injuries (<10) vs Event Weekday (1948-2017)")
```

The distribution looks to be fairly distributed across all days of the week.

## Broad Phase of Flight and Weather Conditions

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, !is.na(Weather.Condition) & !is.na(Broad.Phase.of.Flight)),
       aes(x = Weather.Condition, y = Broad.Phase.of.Flight)) +
  geom_jitter(alpha = 0.2) +
  scale_x_discrete(labels = abbreviate) +
  #scale_y_log10(labels = comma) +
  ggtitle("Broad Phase of Flight vs Weather Conditions (1948-2017)")
```

The relationship between weather condition and flight phase of recorded
accidents/incidents was interesting. It seems there is higher frequency of 
recorded observations for certain combinations of weather and what we could
qualify higher risk phases of flight. For instance, IMC flying conditions while
during "cruise" and the "approach".This is an interesting plot as it gives an 
indication of which phases of flight to be particularly worried about given a 
particular weather condition.

## Broad Phase of Flight and Event Month

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, !is.na(Broad.Phase.of.Flight)),
       aes(x = Event.Month, y = Broad.Phase.of.Flight)) +
  geom_jitter(alpha = 0.2) +
  scale_x_discrete(labels = abbreviate) +
  #scale_y_log10(labels = comma) +
  ggtitle("Broad Phase of Flight vs Event Month (1948-2017)")
```

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, !is.na(Broad.Phase.of.Flight)),
       aes(x = Event.Month, y = Broad.Phase.of.Flight)) +
  geom_bin2d(alpha = 1.0) +
  ggtitle("Broad Phase of Flight vs Weather Conditions (1948-2017)")
```

The plot indicates that there is a higher frequency of recorded observations 
for nothern summer months during Landing and Take-off. Across all months,
the heat map also shows that the Take-off and Landing register the highest 
number of observations.

## Longitude and Latitude 

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, !is.na(Longitude) & !is.na(Latitude) 
              & Latitude < 75 & Latitude > 10 & Longitude < -60),  
       aes(y = Latitude, x = Longitude)) +
  geom_point(shape=21, alpha = 0.2) +
  scale_y_continuous(breaks = seq(-70,100,10)) +
  scale_x_continuous(breaks = seq(-180,180,10)) +
  ggtitle("Latitude vs Longitude (US-focused) (1948-2017)")
```

These plots are essently maps of the US and clearly shows that the coastal 
states are more heavily impacted that the Mid-western states and most of Alaska. 
This can be explained by the volume of flight to/from destinations across the 
US. A sad chart however as it shows that the majority of US States suffered an 
aviation tragedy between 1948 and 2017.

# Multivariate Plots Section
## Longitude and Latitude by Month
First, let's look at Latitude vs Longitude of observations by Month of the Year.

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, !is.na(Longitude) & !is.na(Latitude) 
              & Latitude < 75 & Latitude > 10 & Longitude < -60),  
       aes(y = Latitude, x = Longitude, fill = Event.Month)) +
  geom_point(alpha = 0.3, shape = 21) +
  scale_y_continuous(breaks = seq(-70,100,10)) +
  scale_x_continuous(breaks = seq(-180,180,10)) +
  ggtitle("Latitude vs Longitude (US-focused) by Month (1948-2017)")
```

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, !is.na(Longitude) & !is.na(Latitude) 
              & Latitude < 75 & Latitude > 10 & Longitude < -60),  
       aes(y = Latitude, x = Longitude, fill = Event.Month)) +
  geom_point(alpha = 0.3, shape = 21) +
  facet_wrap(~ Event.Month) +
  scale_y_continuous(breaks = seq(-70,100,20)) +
  scale_x_continuous(breaks = seq(-180,180,50)) +
  ggtitle("Latitude vs Longitude (US-focused) by Month (1948-2017)")
```

The distribution of month across latitude and longitude is fairly spread across 
the US, with a slightly higher prevalence of observations during the winter 
in southern states like Florida.

## Longitude and Latitude by Weather Conditions
Let's now look at latitude vs longitude add layer for the weather condition.

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, !is.na(Longitude) & !is.na(Latitude) 
              & Latitude < 75 & Latitude > 10 & Longitude < -60
              & Weather.Condition != "NA"),  
       aes(y = Latitude, x = Longitude, fill = Weather.Condition)) +
  geom_point(alpha = 0.3, shape = 21) +
  facet_wrap(~ Weather.Condition, ncol = 2) +
  scale_y_continuous(breaks = seq(-70,100,10)) +
  scale_x_continuous(breaks = seq(-180,180,20)) +
  ggtitle("Latitude vs Longitude (US-focused) by Month (1948-2017)")
```

Weather condition VMC seems to be quite consistent except for patches 
of primarily IMC conditions for certain discrete areas.

## Broad Phase of Flight and Event Month by Weather Conditions
Let's now look at the relationship of Broad Phase of Flight vs Month by Weather
Conditions.

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, !is.na(Broad.Phase.of.Flight) 
              & Weather.Condition != "NA" & Weather.Condition != "NA"),
       aes(x = Event.Month, y = Broad.Phase.of.Flight, 
           fill = Weather.Condition)) +
  geom_jitter(shape=21, alpha = 0.3) +
  facet_wrap(~ Weather.Condition, ncol = 2) +
  scale_x_discrete(labels = abbreviate) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  +
  ggtitle("Broad Phase of Flight vs Event Month (1948-2017)")
```

When looking at the relationship between Broad Phase of Flight vs Month and 
Weather Condition, we can see that accident/incidents primarily take place 
during VMC weather condition. However, for certain months of the year such as 
December an January, IMC conditions are a non-negligible portion of the 
observations, particularly during Approach and Cruise phases of flight.

## Broad Phase of Flight and Event Month by Weather Conditions
Next, let's look at Total Fatal Injuries vs Broad Phases of Flight by Weather
Condition.

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, Total.Fatal.Injuries < 40),  
       aes(x = Broad.Phase.of.Flight, y = Total.Fatal.Injuries, fill = Weather.Condition)) +
  geom_jitter(shape=21, alpha = 0.3) +
  scale_x_discrete(labels = abbreviate) +
  scale_y_continuous(breaks = seq(0,40,5)) +
  ggtitle("Total Fatal Injuries (<40) vs Broad Phase of Flight (1948-2017)")
```

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, Total.Fatal.Injuries < 40 
              & Weather.Condition != "UNK" & Weather.Condition != "NA"),  
       aes(x = Broad.Phase.of.Flight, y = Total.Fatal.Injuries, fill = Weather.Condition)) +
  geom_jitter(shape=21, alpha = 0.3) +
  facet_wrap(~ Weather.Condition) +
  scale_x_discrete(labels = abbreviate) +
  scale_y_continuous(breaks = seq(0,40,5)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  +
  ggtitle("Total Fatal Injuries (<40) vs Broad Phase of Flight (1948-2017)")
```

In terms of the relationship between Total Fatal Injuries vs Broad Phases of 
Flight and Weather Condition, the scatter plot clearly indicates that the total
number of fatalies shows an increased number with IMC as weather condition and
during Cruise and Approach phases of flight. Furthermore, we can see that in 
general, low fatality count (less then 2) primarily took place in VMC weather
conditions.

## Total Fatal Injuries and Engine Type by Year
Finally, let's look at the relationship between Total Fatal Injuries vs 
Engine Type by Year. 

```{r message=FALSE, warning=FALSE}
ggplot(subset(pf_study, Total.Fatal.Injuries < 300),  
       aes(y = Total.Fatal.Injuries.Brackets, x = Engine.Type, 
           fill = Event.Year)) +
  geom_jitter(shape = 21, alpha = 0.4) +
  scale_x_discrete(labels = abbreviate) +
  scale_y_discrete(labels = c('1','2','3', '4', '5', 
                              '6-10', '11-20', '21-50', '51-100',
                              '101-150', '151-200', '201-250', '251-300')) +
  scale_fill_gradient(name = "count", trans = "log", 
                      breaks = my_breaks, labels = my_breaks) +  
  ggtitle("Total Fatal Injuriesvs Event Month (1948-2017)")
```

The relationship between Total Fatal Injuries vs Engine Type by Year is 
interesting as it also how certain engines have become prevaent in different 
time periods. For instance, Turbo-Jet and Turbo-Fan powered aircraft show a 
higher number of fatalities in later years whereas Reciprocated engies show a 
distribution of fatalities in earlier years.

# Concluding Remarks

During this EDA, I was particularly suprised by three aspects: 
(1) the amount of time it took to read about and understand each variable to 
make the best use of the given data;
(2) the amount of computational power it took to process a dataset of a little 
less then 80,000 observations - a small sample size in our day & age;
(3) and finally how much data wrangling it took to prep the data despite the 
fact that this is considered "clean data" after Kaggle already did the tidying.

The info from such study on Aviation Accidents could potentially inform pilots 
to pay particular attention for certain combination of weather conditions and 
phases of flight.

### Note
This R Markdown file was created as part of a lesson in UDacity's [Data Analyst Nanodegree](https://ae.udacity.com/course/data-analyst-nanodegree--nd002). 
